---
title: "Untitled"
format: html
editor: visual
---

## Introduction: 
  > The base data set for this workthrough is a binary data set of survey responses to a CDC behavioral and chronic disease surveillance study done for diabetes. I have chosen `mental health`, `healthcare`, `ability/inability to see a physicians due to the cost of the visit` and `difficulty walking as my variables`. The response variable is the `presence of prediabetes or diabetes`. The purpose of this EDA is to get a general idea of how the chosen variables effect levels of diabetes. Ultimately, I will be using these variables to model the likelyhood of diabetes occuring at different levels of variables. 

## Initial Library Read-in:
```{r, warning = FALSE, message = FALSE}
library(readr)
library(readxl)
library(Rmisc)
library(dplyr)
library(tidyverse)
library(ggplot2)
```

## Data:
```{r}
#reading in the csv 
diabetes_data <- read_csv("./data/diabetes_binary_health_indicators_BRFSS2015.csv")
```

```{r}
#checking on missing values
sum(is.na(diabetes_data))
```

```{r}
#changing variables to factor with valid levels; added the if statements to prevent the code from not running when the columns were already changed
if(is.numeric(diabetes_data$Diabetes_binary)) diabetes_data$Diabetes_binary <- cut(diabetes_data$Diabetes_binary, 2, labels = c("No", "Pre/Yes"))

if(is.numeric(diabetes_data$HighBP)) diabetes_data$HighBP <- cut(diabetes_data$HighBP, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$HighChol)) diabetes_data$HighChol <- cut(diabetes_data$HighChol, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$CholCheck)) diabetes_data$CholCheck <- cut(diabetes_data$CholCheck, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$Smoker)) diabetes_data$Smoker <- cut(diabetes_data$Smoker, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$Stroke)) diabetes_data$Stroke <- cut(diabetes_data$Stroke, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$HeartDiseaseorAttack)) diabetes_data$HeartDiseaseorAttack <- cut(diabetes_data$HeartDiseaseorAttack, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$PhysActivity)) diabetes_data$PhysActivity <- cut(diabetes_data$PhysActivity, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$Fruits)) diabetes_data$Fruits <- cut(diabetes_data$Fruits, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$Veggies)) diabetes_data$Veggies <- cut(diabetes_data$Veggies, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$HvyAlcoholConsump)) diabetes_data$HvyAlcoholConsump <- cut(diabetes_data$HvyAlcoholConsump, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$NoDocbcCost)) diabetes_data$NoDocbcCost <- cut(diabetes_data$NoDocbcCost, 2, labels = c("No", "Yes"))                                    
if(is.numeric(diabetes_data$AnyHealthcare)) diabetes_data$AnyHealthcare <- cut(diabetes_data$AnyHealthcare, 2, labels = c("No", "Yes"))    

if(is.numeric(diabetes_data$GenHlth)) diabetes_data$GenHlth <- cut(diabetes_data$GenHlth, 5, labels = c("Excellent", "Very good", "Good", "Fair", "Poor"))

if(is.numeric(diabetes_data$DiffWalk)) diabetes_data$DiffWalk <- cut(diabetes_data$DiffWalk, 2, labels = c("No", "Yes"))

if(is.numeric(diabetes_data$Sex)) diabetes_data$Sex <- cut(diabetes_data$Sex, 2, labels = c("Male", "Female"))

if(is.numeric(diabetes_data$Age)) diabetes_data$Age <- cut(diabetes_data$Age, 13, labels = c("18 to 24", "25 to 29", "30 to 34", "35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64", "65 to 69", "70 to 74", "75 to 79", "80 or older"))

if(is.numeric(diabetes_data$Education)) diabetes_data$Education <- cut(diabetes_data$Education, 6, labels = c("None", "Elementary", "Middle School", "High School", "Some/Technnical College", "College Graduate"))

if(is.numeric(diabetes_data$Income)) diabetes_data$Income <- cut(diabetes_data$Income, 8, labels = c("< $10,000", "< $15,000", "< $20,000", "< $25,000", "< $35,000", "< $50,000", "< $75,000", "> $75,000"))
```

## Summaries:
```{r}
#getting an idea of the count of different levels, this can also provide a general idea of how likely it is for someone to have diabetes under certain conditions; this is further shown through the plots
diabetes_data |>
  group_by(Diabetes_binary, MentHlth, AnyHealthcare, NoDocbcCost, DiffWalk) |>
  tally()
```

```{r}
p1 <- ggplot(diabetes_data, aes(Diabetes_binary, fill = interaction(AnyHealthcare, NoDocbcCost))) + 
      geom_bar(position = "dodge")

p2 <- ggplot(diabetes_data, aes(Diabetes_binary, fill = interaction(DiffWalk, NoDocbcCost))) + 
      geom_bar(position = "dodge")

p3 <- ggplot(diabetes_data, aes(Diabetes_binary, fill = interaction(AnyHealthcare, DiffWalk))) + 
      geom_bar(position = "dodge")

multiplot(p1,p2,p3)
```

```{r}
ggplot(diabetes_data, aes(x = Diabetes_binary, fill = AnyHealthcare)) +
    geom_bar(position = "dodge") +
    facet_grid(NoDocbcCost~DiffWalk, labeller=label_both) 
```

